Validation of Visual Statistical Inference, with Application to Linear Models

### 2012-04-23 paper_jasa.tex
model with subject specific ability

The scientific foundation of graphical methods for data analysis is well established by \cite{cleveland:1984}. In recent years we have seen several major advances in statistical graphics. Modern computing systems like R and SAS tease the production of high quality statistical plots. A grammar of graphics introduced by \cite{wilkinson:1999} presents a structured way to generate specific graphics from data and define connections between disparate types of plots.  \cite{hadley:2009} has implemented a revised version of the grammar of graphics in R, in the package {\tt ggplot2}.   \blue {Graham Wills has implemented the grammar in SPSS, but I'm not sure how to cite this.}

@@

\section{Visual Statistical Inference} \label{sec:visual_test}

@@

When $K$ individuals evaluate a lineup plot independently, the number of successful evaluations, $U \sim \text{Binom} (K,1/m)$  under $H_0$. This leads us to obtain the probability $$Pr(U \ge u)= \sum_{k \ge u}^K {{K \choose k} \left(\frac{1}{m}\right)^k\left(1-\frac 1m\right)^{(K-k)}}$$ where $u$ is the observed number of successful evaluations. We use this probability to  define the $p$-value for visual test $T_{\theta}$ as follows:
\begin{dfn} \label{dfn:pval}
The $p$-value associated with the decision of a visual test $T_{\theta}$ is given \blue{as the probability to observe a value at least as extreme as the test statistic $u$, 
where $u$ is the number of successful evaluations of the lineup, which leads us to a $p$-value of $P(U \ge u)$. Since $u$ is a discrete value with boundaries 0 and $K$,
 we can also re-formulate the $p$-value in terms of upper and lower boundaries:
 \begin{equation*}
 \begin{array}{rcll}
 P(U \ge u+1) \le &p\text{-value}& (\le 1), & \text{if }u < K, \text{and in particular for } u=1\\
( 0 \le) &p\text{-value}& \le P(U \ge u-1), & \text{if }u > 1, \text{and in particular for } u=K.\\ 
 \end{array}
 \end{equation*}
 }
\red{HH: Tried to re-write it to answer the question below.
Why is the $p$ value not just $Pr(U \ge u)$ ? (We need to modify this a bit with respect to the rejection region $R$, but there should not be a difference in terms of whether we reject or accept. }

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=3.5in]{images/p_val_power_m.pdf} 
   \caption{Comparison of probabilities to pick the data plot for different size lineups $m$ and different values for data plot strength $p_D$. The grey line shows probability of picking the data plot under $H_0$. }
   \label{fig:pval_power}
\end{figure}

Now assume that the same lineup plot is evaluated by $K$ independent observers. W.l.o.g. let the last plot be the data plot and plots 1 through $m-1$ the null plots. Let $n_i$ be the number of observers who picked plot $i$ as the plot with the structure least consistent with the null hypothesis, then $\sum_{k=1}^{m} n_k = K$, and $(n_1, n_2, ..., n_m) \sim$ Mult$_{\pi_1, ..., \pi_m}$ with $\sum_i \pi_i = 1$. We can estimate $\pi_i$ as $\widehat{\pi_i} = n_i/K$.

For the distribution of $n_m$, %the number of times out of $K$ that observers picked the last plot to show the structure least consistent with the null hypothesis, 
we get
\[
n_m\sim \left \{ 
\begin{array}{ll}
B_{K, (1-E[p_D])^{m-1}} & \text { under } H_1\\
B_{K, 1/m} & \text { under } H_0
\end{array}
\right .
\]

Under $H_1$ we therefore get an estimate for the strength of the signal in the test statistic based on a lineup evaluation as 
\begin{equation}\label{eqn:power_estimate}
\widehat{E[p_D]} = 1- \left(\frac{n_m}{K}\right)^{1/(m-1)}.
\end{equation}

\blue{Note that the right hand side of equation (\ref{eqn:power_estimate}) is independent of the parameter $\theta$, but just based on the lineup evaluation by independent observers. This allows us to compute this value independently of the parametric setting. We therefore define the {\it signal strength} of a  visual test as the following:

\begin{dfn} \label{dfn:signal}
The {\it signal strength} $\sigma_T$ of a visual test is defined as the probability to identify the data plot.
\end{dfn}

In a lineup of size $m$, evaluated by $K$ independent observers,  we can estimate the signal strength of the test as
\[
\sigma_T^{m-1} = \frac{n_m}{K}.
\]

In a parametric setting, signal strength of a test coincides with the power of the test.
}

\red{We need to comment on how this ties in to the rest of the paper: this is a theoretical consideration based on the assumption that observers are able to spot the plot with the smallest/a small $p$-value.}

>>>>

The scientific foundation of graphical methods for data analysis is well established by \cite{cleveland:1984}. In recent years we have seen several major advances in statistical graphics. Modern computing systems like R and SAS ease the production of high quality statistical plots. A grammar of graphics introduced by \cite{wilkinson:1999} presents a structured way to generate specific graphics from data and define connections between disparate types of plots.  \cite{hadley:2009} has implemented a revised version of the grammar of graphics in R, in the package {\tt ggplot2}.   \blue {Graham Wills has implemented the grammar in SPSS, but I'm not sure how to cite this.}

@@

\section{Visual Statistical Inference} \label{sec:visual_test} 
Visual tests are non-parametric tests of the form $H_0$: the data plot is visually not distinguishable from a null plot vs $H_1$: the data plot is visually distinguishable from a plot generated consistently with $H_0$.

Power of a plot and $p$-value have been derived in Buja and Wickham as ...

for this paper, we want to place the parametric inference of a regression setting into the framework of visual testing:

@@

When $K$ individuals evaluate a lineup plot independently, the number of successful evaluations, $U \sim \text{Binom} (K,1/m)$  under $H_0$. This leads us to obtain the probability 
\begin{equation}\label{eqn:pvalue}
Pr(U \ge u)= \sum_{k \ge u}^K {{K \choose k} \left(\frac{1}{m}\right)^k\left(1-\frac 1m\right)^{(K-k)}}
\end{equation}
where $u$ is the observed number of successful evaluations. We use this probability to  define the $p$-value for visual test $T_{\theta}$ as follows:
\begin{dfn} \label{dfn:pval}
The $p$-value associated with the decision of a visual test $T_{\theta}$ is given \blue{as the probability to observe a value at least as extreme as the test statistic $u$, 
where $u$ is the number of successful evaluations of the lineup, which leads us to a $p$-value of $P(U \ge u)$ as given in eqn (\ref{eqn:pvalue}). Since $u$ is a discrete value with boundaries 0 and $K$,
 we can also re-formulate the $p$-value in terms of upper and lower boundaries:
 \begin{equation*}
 \begin{array}{rcll}
 P(U \ge u+1) \le &p\text{-value}& (\le 1), & \text{if }u < K, \text{and in particular for } u=0\\
( 0 \le) &p\text{-value}& \le P(U \ge u-1), & \text{if }u > 1, \text{and in particular for } u=K.\\ 
 \end{array}
 \end{equation*}
 }
\red{HH: Tried to re-write it to answer the question below.
Why is the $p$ value not just $Pr(U \ge u)$ ? (We need to modify this a bit with respect to the rejection region $R$, but there should not be a difference in terms of whether we reject or accept. }

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=3.5in]{images/powerplot.pdf} 
   \caption{Comparison of probabilities to pick the data plot for different size lineups $m$ and different values for data plot strength $p_D$. The grey line shows probability of picking the data plot under $H_0$. }
   \label{fig:pval_power}
\end{figure}

Now assume that the same lineup plot is evaluated by $K$ independent observers. W.l.o.g. let the last plot be the data plot and plots 1 through $m-1$ the null plots. Let $n_i$ be the number of observers who picked plot $i$ as the plot with the structure least consistent with the null hypothesis, corresponding to random variable $N_i$.  Then  $(N_1, N_2, ..., N_m) \sim$ Mult$_{\pi_1, ..., \pi_m}$ with $\sum_i \pi_i = 1$ and $\sum_{k=1}^{m} n_k = K$. We can estimate $\pi_i$ as $\widehat{\pi_i} = n_i/K$.

For the distribution of $N_m$, %the number of times out of $K$ that observers picked the last plot to show the structure least consistent with the null hypothesis, 
we get
\[
N_m\sim \left \{ 
\begin{array}{ll}
B_{K, (1-E[p_D])^{m-1}} & \text { under } H_1\\
B_{K, 1/m} & \text { under } H_0
\end{array}
\right .
\]

By matching the expected values, we therefore get an expression for the $p$-value in the data under $H_1$  as:
\begin{equation}\label{eqn:power_estimate}
{E[p_D]} = 1- \left(\frac{E[N_m]}{K}\right)^{1/(m-1)}.
\end{equation}

\blue{Note that the right hand side of equation (\ref{eqn:power_estimate}) is independent of the parameter $\theta$, but just based on the lineup evaluation by independent observers. While this allows us to compute this value independently of the parametric setting, it also means that onlookers might pick out the data plot for a reason not related to the value of $\theta$.

%Picking out the data plot is still a measure of how different the data plot is compared to the null plots, and therefore is a test statistic of the visual test.

%This allows us to compute this value independently of the parametric setting. We therefore define the {\it signal strength} of a  visual test as the following:
%
%\begin{dfn} \label{dfn:signal}
%The {\it signal strength} $\sigma_T$ of a visual test is defined as the probability to identify the data plot.
%\end{dfn}
%
%In a lineup of size $m$, evaluated by $K$ independent observers,  we can estimate the signal strength of the test as
%\[
%\sigma_T^{m-1} = \frac{n_m}{K}.
%\]
%
%In a parametric setting, signal strength of a test coincides with the power of the test.
}

\red{We need to comment on how this ties in to the rest of the paper: this is a theoretical consideration based on the assumption that observers are able to spot the plot with the smallest/a small $p$-value. $p$ value of data plot and $p$ values from null are negatively correlated (if one is low, the other has to be higher. We need to address this.}
\red{HH: should we re-run some of the lineups with different null plots?}

\blue{
\paragraph{Subject-specific Power}
Probability to reject null hypothesis is power. This is intricately related to the probability to identify the data plot from the lineup. We need to address the problematic of individual specific abilities in identifying the data plot
}. 
\red{HH: need to cite some literature from cognitive science here}

\blue{Let $X_{j\ell} \sim B_{1, \pi_{j\ell}}$ be the random variable, with $X_{j\ell}= 1$, if  subject $j$ correctly identifies the data plot from lineup $\ell$, and 0 otherwise.
We can model $\pi_{j\ell}$ with a logistic regression with subject specific random effects as:
\[
\text{logit } \pi_{j\ell} = X_\ell \beta + Z u_j,  
\]
where $X_\ell$ is a matrix of fixed effects with parameter vector $\beta$ corresponding to lineup $\ell$ and $u \sim MVN(0, \sigma_u I)$  is a vector of random variables $j = 1, ..., J$.
}
